---
title: "Module 3: Hospital Pricing and Competition"
subtitle: "Part 0: Motivation"
author: Ian McCarthy | Emory University
date: Econ 372 #"`r format(Sys.time(), '%d %B %Y')`"
header-includes: 
  - \usepackage{tikz}
  - \usepackage{amsmath}
  - \usepackage{pgf}
  - \usepackage[xcolor=pst]{pstricks}
  - \usepackage{pgfshade}
  - \usepackage{pdftools}
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "macros.js"
knit: pagedown::chrome_print
---

<!-- Adjust some CSS code for font size and maintain R code font size -->
<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 2em 1em 2em;    
}
.remark-code, .remark-inline-code { 
    font-size: 20px;
}
</style>


<!-- Set R options for how code chunks are displayed and load packages -->
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=3, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )

knitr::opts_hooks$set(fig.callout = function(options) {
  if(options$fig.callout) {
    options$echo = FALSE
  }
  options
})

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes,
               scales, gganimate, gapminder, gifski, png, tufte, plotly, OECD,
               ggrepel, survey, foreign, devtools, pdftools, xaringanExtra)
```

```{r xaringan-scribble, echo=FALSE}
xaringanExtra::use_scribble(rgb(0.9, 0.5, 0.5))
```


# History of hospitals

- Before 1900: Just don't go to the hospital! (at least in the U.S.)
  - mainly charity care
  - hospitals were a learning experience for physicians

- Early 1900s: big safety and technological improvements

- Mid 1900s: huge growth, especially in wealthy and urban areas
  - Medicare and Medicaid in 1965 (Social Security Act)

---
# Now
As we know, we now spend <b>a lot</b> on health care in the U.S., and a big part of that is very high health care prices (and a big part of that is hospital prices)


---
# Ownership types

1. Private not-for-profit: About 60%

2. For-profit: About 20%

3. State and local gov't: About 20%

<br>
<div class="smalltext">Source: <a href="https://www.aha.org/statistics/fast-facts-us-hospitals">AHA Fast Facts</a></div>

---
# Non-profit hospitals
What does it mean to be a not-for-profit hospital?<br>

--

From an economics perspective:
- Hospital assumed to maximize some objective function, $u(q,z)$, subject to a production constraint

- $q$ denotes quantity of care and $z$ denotes quality of care

- Production is constrained by the break-even condition

---
# Non-profit hospitals
What does it mean to be a not-for-profit hospital?<br>

--

From a practical perspective:
- Profits must be re-invested into the hospital

- Must show "community benefit" (no consensus definition...includes uncompensated care, services to Medicaid, and certain specialized services that are generally unprofitable)

- No taxes! and tax-free bonds

---
# Non-profit hospitals and tax benefits
- <span>&#36;</span>24.6 billion in tax exemptions in 2011

- <span>&#36;</span>62.4 billion in "community benefits"<br>

- [Washington Post Article](https://www.washingtonpost.com/national/health-science/value-of-tax-breaks-for-nonprofit-hospitals-doubled-in-a-decade/2015/06/17/4162c640-1450-11e5-9ddc-e3353542100c_story.html?utm_term=.4f877a72a09f)<br>

--

What do you think? Are these community benefits measured appropriately?


---
# What is a non-profit hospital?
The real question is...what is the hospital's objective function?

- For-profit in disguise
- Output maximizers
- Tax-benefit maximizers
- Social welfare maximizers<br>

--

Most empirical evidence doesn't find much of a difference between FP and NFP hospitals, except FPs have higher prices. Why is that?


---
# For-profit hospitals
These are easier to study theoretically...just a standard profit maximizing firm.

- $\pi=P(q)q - C(q),$ where $q$ denotes quantity of care

- Firm has some market power and so faces a downward sloping demand curve


