\documentclass[11pt,]{article}
\usepackage[margin=1in]{geometry}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage[]{mathpazo}
\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center
\newcommand{\blankline}{\quad\pagebreak[2]}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}} 
\usepackage{longtable,booktabs}

\usepackage{parskip}
\usepackage{titlesec}
\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}

\titleformat*{\subsubsection}{\normalsize\itshape}

\usepackage{titling}
\setlength{\droptitle}{-.25cm}

%\setlength{\parindent}{0pt}
%\setlength{\parskip}{6pt plus 2pt minus 1pt}
%\setlength{\emergencystretch}{3em}  % prevent overfull lines 

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{lastpage}
\renewcommand{\headrulewidth}{0.3pt}
\renewcommand{\footrulewidth}{0.0pt} 
\lhead{}
\chead{}
\rhead{\footnotesize Econ 372: Economics of Health Care
Markets -- 1/10/2023 - 4/24/2023}
\lfoot{}
\cfoot{\small \thepage/\pageref*{LastPage}}
\rfoot{}

\fancypagestyle{firststyle}
{
\renewcommand{\headrulewidth}{0pt}%
   \fancyhf{}
   \fancyfoot[C]{\small \thepage/\pageref*{LastPage}}
}

%\def\labelitemi{--}
%\usepackage{enumitem}
%\setitemize[0]{leftmargin=25pt}
%\setenumerate[0]{leftmargin=25pt}




\makeatletter
\@ifpackageloaded{hyperref}{}{%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
}
\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={ ()},
             pdfkeywords = {},  
            pdftitle={Econ 372: Economics of Health Care Markets},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls


\setcounter{secnumdepth}{0}

\usepackage{longtable}




\usepackage{setspace}

\title{Econ 372: Economics of Health Care Markets}
\author{Ian McCarthy}
\date{1/10/2023 - 4/24/2023}


\begin{document}  

		\maketitle
		
	
		\thispagestyle{firststyle}

%	\thispagestyle{empty}


	\noindent \begin{tabular*}{\textwidth}{ @{\extracolsep{\fill}} lr @{\extracolsep{\fill}}}


E-mail: \texttt{\href{mailto:ian.mccarthy@emory.edu}{\nolinkurl{ian.mccarthy@emory.edu}}} & Web: \href{http://econ372s22.classes.ianmccarthyecon.com}{\tt econ372s22.classes.ianmccarthyecon.com}\\
Office Hours: MW 1:30-2:30 p.m.  &  Class Hours: MW 11:30-12:45 p.m.\\
Office: R. Rollins, R432  & Class Room: White Hall 103\\
	&  \\
	\hline
	\end{tabular*}
	
\vspace{2mm}
	


\hypertarget{course-description}{%
\section{Course Description}\label{course-description}}

This course examines the industrial organization of health care markets
in the U.S., including underlying theory, empirical findings, and
related health care policy. We will focus specifically on insurance
markets, informational asymmetries between physicians and patients, and
hospital pricing.

\textbf{Prerequisites:} Economics 101 (Principles of Microeconomics).
Some basic knowledge of calculus and derivatives is also required, but
not necessarily the official Math 111 (Calculus I) course at Emory.

\hypertarget{learning-outcomes}{%
\section{Learning Outcomes}\label{learning-outcomes}}

The broad goal of the class is to understand the structure of health
care markets in the U.S. The course is designed around three areas,
reflecting the different interactions an individual would have when
navigating the health care system (choosing an insurance plan, visiting
a physician, and receiving hospital care). For each segment, we will
develop the underlying theoretical model(s), cover recent empirical
studies in the area, and discuss relevant health care policy. My
specific goals are that, by the end of this course, you will be able to:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Explain the structure of the U.S. health care system, its main
  components, and its history;
\item
  Model adverse selection in health insurance, examine its effects on
  health insurance markets, and support your arguments with existing
  data;
\item
  Explain the physician agency problem, use a model of physician agency
  to examine financial incentives in health care, and summarize
  empirical evidence on the presence of physician agency; and
\item
  Describe hospital pricing, negotiation with insurers, and explain
  (qualitatively and quantitatively) the differences between charges and
  prices.
\item
  Analyze hospital data in a real-life setting and predict effects of
  real-life policies
\end{enumerate}

There are more specific learning outcomes for each module described on
the relevant module's page of our class website, accessible
\href{https://econ372s23.classes.ianmccarthyecon.com/}{here}.

\hypertarget{text-software-and-class-materials}{%
\section{Text, Software, and Class
Materials}\label{text-software-and-class-materials}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Where to go:} The best place for information on the class is
  our class website,
  \href{https://econ372s23.classes.ianmccarthyecon.com/}{here}. I use
  the website as an easier way for you to access course materials,
  including notes/slides, a detailed schedule, and assignments. I'll
  also use Canvas to distribute any papers or specific readings that
  aren't available publicly, as well as post information that I don't
  want public like virtual meeting passwords (if needed) and grades.
\item
  \textbf{Readings:} There are two optional textbooks for this class.
  One is an intermediate microeconomics textbook that will cover much of
  the basic economic theory, and another is a health economics textbook.
  We will supplement these textbooks with other readings listed below in
  the detailed schedule, as well as online. These textbooks are not
  required but may be useful for students looking for some supplemental
  resources.

  \begin{itemize}
  \tightlist
  \item
    R. Pindyck and D. Rubinfeld \emph{Microeconomics} (Upper Saddle
    River, NJ: Prentice Hall, 2018).
  \item
    Frank A Sloan and Chee-Ruey Hsieh \emph{Health Economics}, vol. 1
    (The MIT Press, 2012).
  \end{itemize}
\item
  \textbf{Statistics Software:} Each assignment will generally include
  both a theoretical and empirical section. Most of the empirical work
  involves basic summary statistics. This can be done in just about any
  spreadsheet or statistics package. I'll use \texttt{R} for my work,
  but you are free to use whatever software you're most comfortable
  with. I encourage you to use \texttt{R} or \texttt{Python} simply
  because these are the most common programs used in applied data
  analysis right now, but you can certainly use Excel or Google Sheets
  if you'd prefer. This is \textbf{not} a statistics or econometrics
  class, so I do not expect you to have any statistics programming
  knowledge.
\item
  \textbf{Accessing Data:} For most empirical work, I will house all of
  the links to the data on our class website. For transparency, I will
  also point you to other GitHub repositories where available. If you're
  new to Git or GitHub and want to understand some of that better, take
  a look at \href{https://grantmcdermott.com/}{Grant McDermott's} notes
  on \href{https://github.com/uo-ec607/lectures}{Data Science for
  Economists} as well as \href{https://jennybryan.org/}{Jenny Bryan's}
  online reference book, \href{https://happygitwithr.com/}{Happy Git and
  GitHub with R}. But again, knowledge of Git or GitHub is not part of
  this class.
\item
  \textbf{Turning things in:} All assignments should be submitted as PDF
  files in Canvas. Since we'll be using some math in all of our answers,
  I suggest that you try \href{https://www.overleaf.com/}{Overleaf}, but
  you can also write your answers in Word and print to PDF. As an Emory
  student, everyone has access to an institutional Overleaf account,
  which provides you with all the nice features of the product. You can
  use this to easily write PDF documents in a language known as
  ``LaTeX''. For a brief introductory video and links to more info,
  please see the video
  \href{https://www.overleaf.com/learn/latex/Questions/Getting_started_with_LaTeX_and_Overleaf_(formerly_writeLaTeX)}{here}.
\end{enumerate}

\hypertarget{course-policies}{%
\section{Course Policies}\label{course-policies}}

Various policies for this course are described below. Basically, let's
all work to be good citizens and take seriously our various roles as a
student, teacher, friend, colleague, human, etc.

\hypertarget{class-meetings}{%
\subsection{Class meetings}\label{class-meetings}}

All regular class meetings will take place in \textbf{White Hall, Room
103} on Mondays and Wednesdays from 11:30am to 12:45pm.

\hypertarget{office-hours}{%
\subsection{Office hours}\label{office-hours}}

My designated office hours will be on Monday and Wednesday from 1:30 to
2:30pm in R. Rollins, R432. I'm also available outside of these times. I
just ask that you schedule an appointment in advance.
\href{https://mccarthy-meetings.youcanbook.me/}{Make an appointment with
me here.}

While we call these ``office hours'' they really should be ``student
hours''. These are set times dedicated to all of you. This means that I
will be available, waiting for you to come by with whatever questions
you have. This is the easiest way to find me and usually the best chance
for discussing class material and concerns.

Finally, for any questions that you don't think need a meeting, you can
always reach out to me through email. I do my best to respond within 24
hours to all emails.

\hypertarget{ta}{%
\subsection{TA}\label{ta}}

Our TA this semester is Rachel Clohan. She's a PhD student in the
economics department, interested in research related to health and
health care economics. Rachel is awesome and I'm sure will be a great
resource for you throughout the semester. She will attend most (but not
all) of the classes and will hold office hours on \textbf{Thursdays}
from \textbf{3:00pm to 4:00pm} in the R. Rollins Building, R400C-25. She
has a cubicle just to the right of the entrance into the fourth floor
offices. You can also reach Rachel via email at
\href{mailto:rachel.clohan@emory.edu}{\nolinkurl{rachel.clohan@emory.edu}}.

\hypertarget{academic-integrity}{%
\subsection{Academic integrity}\label{academic-integrity}}

The Emory University Honor Code is taken seriously and governs all work
in this course. Details about the Honor Code are available
\href{http://catalog.college.emory.edu/academic/policies-regulations/honor-code.html}{here}.
By taking this course, you affirm that it is a violation of the code to
cheat on exams, to plagiarize, to deviate from the instructions about
collaboration on work that is submitted for grades, to give false
information to a faculty member, and to undertake any other form of
academic misconduct. You agree that the teacher is entitled to move you
to another seat during examinations, without explanation. You also
affirm that if you witness others violating the code you have a duty to
report them to the honor council.

\hypertarget{accessibility-services}{%
\subsection{Accessibility services}\label{accessibility-services}}

If you anticipate issues related to the format or requirements of this
course, please meet with me. I would like us to discuss ways to ensure
your full participation in the course. If you determine that
accommodations are necessary, you may register with Accessibility
Services at (404)727-9877 or via e-mail at
\href{mailto:accessibility@emory.edu}{\nolinkurl{accessibility@emory.edu}}.
To register with OAS, students must self identify and initiate contact
with the OAS office.

\hypertarget{class-wide-announcements}{%
\subsection{Class-wide announcements}\label{class-wide-announcements}}

I will post regular announcements to the class on \textbf{Canvas}, so
please set up your notifications accordingly. I will also use Canvas to
post all grades and any other information that needs to stay in the
class. All other course materials will be available on our class
website, \href{https://econ372s23.classes.ianmccarthyecon.com/}{here}.

\hypertarget{laurens-promise}{%
\subsection{Lauren's Promise}\label{laurens-promise}}

Lauren McCluskey, a 21-year-old honors student athlete,
\href{https://www.sltrib.com/opinion/commentary/2019/02/10/commentary-failing-lauren/}{was
murdered on October 22, 2018 by a man she briefly dated on the
University of Utah campus}. We must all take action to ensure that this
never happens again. If you are in immediate danger, call 911 or Emory
police (404-727-6111). \textbf{I will listen and believe you if someone
is threatening you.}

Any form of sexual harassment or violence will not be excused or
tolerated at Emory. If you are experiencing sexual assault, domestic
violence, or stalking, please report it to me or directly to
\href{http://respect.emory.edu/index.html}{Emory's Office of Respect}
(470-270-5360).

\hypertarget{assignments-and-grading}{%
\section{Assignments and Grading}\label{assignments-and-grading}}

Your final grade consists of three homework assignments, a mid-term
exam, and a class project, along with a participation component. I
describe each assignment below, with more additional detail provided on
our \href{https://econ372s23.classes.ianmccarthyecon.com/}{class
website}.

\hypertarget{participation}{%
\subsection{Participation}\label{participation}}

``Participation'' will be graded based on how many of the individual
participation elements are submitted. Throughout the semester, I will
post ``check-in'' questionnaires on Canvas. I anticipate 6 check-ins
total for the semester. Each check-in will ask questions about what
you've learned in a given module so far, areas where you are confused,
things that we do in class that you think are working, and things that
we do in class that you think are not working. Completion of the
check-in will provide 1 participation point.

We will also have regular in-class exercises. You can think of these as
practice problems that we work on collectively as a class. Completing
these exercises is also worth 1 participation point.

At the end of the semester, I will tally your participation points, and
your final participation grade will be out of 20 possible participation
points. We'll have well over 20 total participation points available, so
you have some margin for error here.

\hypertarget{homework-assignments}{%
\subsection{Homework assignments}\label{homework-assignments}}

There will be 3 homework assignments throughout the semester (one for
each module). These assignments will consist of two parts: 1) short
answer questions related to models covered in class or readings; and 2)
empirical questions related to data I provide or that you are asked to
collect. Preliminary due dates for these assignments are below, and
details of the assignments are on our
\href{https://econ372s23.classes.ianmccarthyecon.com/}{class website}.

\begin{itemize}
\tightlist
\item
  Friday, 2/10
\item
  Friday, 3/17
\item
  Friday, 4/21
\end{itemize}

All assignments are due (as PDFs, submitted via Canvas) by 11:59pm on
the due date. I'll do my best to have all homework assignments graded by
Wednesday of the week after they are submitted.

\hypertarget{exam}{%
\subsection{Exam}\label{exam}}

There is one mid-term exam in this class and no final. The mid-term is
scheduled for \textbf{Wednesday, March 22} and will cover all material
from Modules 1 and 2. Please notify me immediately if you will need
other accommodations or if you already know of an unavoidable scheduling
conflict. In general, I expect everyone to take the exam at the
scheduled time. Make-up exams are a privilege, not a right, and last
minute accommodations will only be made when absolutely necessary.

\hypertarget{final-project}{%
\subsection{Final project}\label{final-project}}

Each of you will prepare a 2-page policy brief, split up into a few
distinct stages. Details of the project are available on the
\href{https://econ372s23.classes.ianmccarthyecon.com/}{class website},
and your final policy brief is due on the last day of class
(\textbf{April 24, 2023}).

The project is an important part of the class and reflects one of the
core objectives --- to analyze data in a real-life setting and apply
those findings to a real-life policy situation. The project is intended
to be a semester-long endeavor and should be the predominant focus
throughout the second half of the course. This is why we don't have a
final exam.

\hypertarget{due-dates}{%
\subsection{Due dates}\label{due-dates}}

This section is just to highlight the most important dates on which an
assignment is due. Late assignments will receive an automatic 10\%
reduction in the grade for each day the assignment is turned in after
the due date.

\begin{itemize}
\tightlist
\item
  \textbf{Friday, 2/10}: Homework 1
\item
  \textbf{Friday, 3/17}: Homework 2
\item
  \textbf{Wednesday, 3/22}: Mid-term exam
\item
  \textbf{Friday, 4/21}: Homework 3
\item
  \textbf{Monday, 4/24}: Final project
\end{itemize}

\hypertarget{grades}{%
\subsection{Grades}\label{grades}}

Each assignment will contribute to your final grade as described below.

\begin{longtable}[]{@{}rll@{}}
\toprule()
Points & Assignment & Percent \\
\midrule()
\endhead
20 & Participation & 5\% \\
180 & Homework (60 x 3) & 45\% \\
60 & Mid-term exam & 15\% \\
140 & Final project & 35\% \\
\bottomrule()
\end{longtable}

Your final percentage grade comes from your total points as a percent of
all possible points available in the class (400). That percent then
translates to a letter grade as follows:

\begin{longtable}[]{@{}llll@{}}
\toprule()
Grade & Range & Grade & Range \\
\midrule()
\endhead
A & 93-100\% & C & 73-76\% \\
A- & 90-92\% & C- & 70-72\% \\
B+ & 87-89\% & D+ & 67-69\% \\
B & 83-86\% & D & 63-66\% \\
B- & 80-82\% & D- & 60-62\% \\
C+ & 77-79\% & F & \textless{} 60\% \\
\bottomrule()
\end{longtable}

\hypertarget{class-schedule}{%
\section{Class Schedule}\label{class-schedule}}

Below is a preliminary outline (subject to change) of specific topics
and assignments throughout the semester.

\hypertarget{module-0-kicking-things-off}{%
\subsection{Module 0: Kicking things
off!}\label{module-0-kicking-things-off}}

\hypertarget{introduction-to-the-economics-of-health-care}{%
\subsubsection{1/11: Introduction to the economics of health
care}\label{introduction-to-the-economics-of-health-care}}

\begin{itemize}
\tightlist
\item
  Kenneth J. Arrow {``Uncertainty and the {Welfare} {Economics} of
  {Medical} {Care},''} \emph{The American Economics Review} 53, no. 5
  (1963), \url{https://www.aeaweb.org/aer/top20/53.5.941-973.pdf}.
\item
  M Gaynor, K Ho, and R Town {``The {Industrial} {Organization} of
  {Health} {Care} {Markets},''} \emph{Journal of Economic Literature}
  47, no. 2 (2015): 235--284.
\end{itemize}

\hypertarget{module-1-health-insurance}{%
\subsection{Module 1: Health
insurance}\label{module-1-health-insurance}}

\hypertarget{understanding-risk}{%
\subsubsection{1/18: Understanding risk}\label{understanding-risk}}

\begin{itemize}
\tightlist
\item
  Sections 5.1-5.2 of Pindyck and Rubinfeld \emph{Microeconomics}.
\end{itemize}

\hypertarget{demand-for-health-insurance}{%
\subsubsection{1/23, 1/25: Demand for health
insurance}\label{demand-for-health-insurance}}

\begin{itemize}
\tightlist
\item
  Sections 4.1-4.7 of Sloan and Hsieh \emph{Health Economics}.
\end{itemize}

\hypertarget{adverse-selection}{%
\subsubsection{1/30, 2/1: Adverse Selection}\label{adverse-selection}}

\begin{itemize}
\tightlist
\item
  Section 17.1 of Pindyck and Rubinfeld \emph{Microeconomics}.
\item
  Sections 4.8-4.10 of Sloan and Hsieh \emph{Health Economics}.
\item
  Liran Einav and Amy Finkelstein {``Selection in {Insurance} {Markets}:
  {Theory} and {Empirics} in {Pictures},''} \emph{The Journal of
  Economic Perspectives} 25, no. 1 (2011): 115--138.
\end{itemize}

\hypertarget{health-insurance-and-current-policy}{%
\subsubsection{2/6, 2/8: Health insurance and current
policy}\label{health-insurance-and-current-policy}}

\begin{itemize}
\tightlist
\item
  Homework 1 due on \textbf{Friday, 2/10}
\end{itemize}

\hypertarget{module-2-physician-agency-and-treatment-decisions}{%
\subsection{Module 2: Physician agency and treatment
decisions}\label{module-2-physician-agency-and-treatment-decisions}}

\hypertarget{physician-agency}{%
\subsubsection{2/13, 2/15: Physician agency}\label{physician-agency}}

\begin{itemize}
\tightlist
\item
  Section 17.4 of Pindyck and Rubinfeld \emph{Microeconomics}.
\item
  Thomas G McGuire {``Physician Agency,''} \emph{Handbook of Health
  Economics} 1 (2000): 461--536.
\end{itemize}

\hypertarget{fee-for-service-payments}{%
\subsubsection{2/20, 2/22: Fee-for-service
payments}\label{fee-for-service-payments}}

\begin{itemize}
\tightlist
\item
  McGuire {``Physician Agency.''}
\item
  \href{https://www.macpac.gov/medicaid-101/}{Medicaid pricing primer}
\item
  \href{https://fas.org/sgp/crs/misc/R40425.pdf}{Medicare pricing
  primer}
\end{itemize}

\hypertarget{capitated-payments}{%
\subsubsection{2/27, 3/1: Capitated payments}\label{capitated-payments}}

\begin{itemize}
\tightlist
\item
  Section 5.3 of Sloan and Hsieh \emph{Health Economics}.
\item
  Jonathan Gruber and Maria Owings {``Physician Financial Incentives and
  Cesarean Section Delivery,''} \emph{The RAND Journal of Economics} 27,
  no. 1 (1996): 99--123.
\item
  Jeffrey Clemens and Joshua D Gottlieb {``Do {Physicians}' {Financial}
  {Incentives} {Affect} {Medical} {Treatment} and {Patient} {Health}?''}
  \emph{American Economic Review} 104, no. 4 (2014): 1320--1349.
\end{itemize}

\hypertarget{current-issues-in-health-care-payments}{%
\subsubsection{3/13, 3/15: Current issues in health care
payments}\label{current-issues-in-health-care-payments}}

\begin{itemize}
\tightlist
\item
  Section 17.2 of Sloan and Hsieh \emph{Health Economics}.
\item
  Homework 2 due \textbf{Friday, 3/17}
\end{itemize}

\hypertarget{mid-term-review-and-exam}{%
\subsubsection{3/20, 3/22: Mid-term Review and
Exam}\label{mid-term-review-and-exam}}

\hypertarget{module-3-hospital-pricing-and-competition}{%
\subsection{Module 3: Hospital pricing and
competition}\label{module-3-hospital-pricing-and-competition}}

\hypertarget{pricing-in-two-price-market}{%
\subsubsection{3/27, 3/29: Pricing in two-price
market}\label{pricing-in-two-price-market}}

\begin{itemize}
\tightlist
\item
  Section 5.4 of Sloan and Hsieh \emph{Health Economics}.
\end{itemize}

\hypertarget{prices-and-bargaining}{%
\subsubsection{4/3, 4/5: Prices and
bargaining}\label{prices-and-bargaining}}

\begin{itemize}
\tightlist
\item
  Uwe E Reinhardt {``The Pricing of {US} Hospital Services: Chaos Behind
  a Veil of Secrecy,''} \emph{Health Affairs} 25, no. 1 (2006): 57--69.
\end{itemize}

\hypertarget{hospital-competition}{%
\subsubsection{4/10, 4/12: Hospital
competition}\label{hospital-competition}}

\begin{itemize}
\tightlist
\item
  Gaynor, Ho, and Town {``The {Industrial} {Organization} of {Health}
  {Care} {Markets}.''}
\item
  M Gaynor, F Mostashari, and P Ginsburg \emph{Making Health Care
  Markets Work: {Competition} Policy for Health Care}, Policy \{Report\}
  (Brookings Institute, 2017).
\end{itemize}

\hypertarget{mergers-prices-and-quality}{%
\subsubsection{4/17, 4/19: Mergers, prices, and
quality}\label{mergers-prices-and-quality}}

\begin{itemize}
\tightlist
\item
  Gaynor, Ho, and Town {``The {Industrial} {Organization} of {Health}
  {Care} {Markets}.''}
\item
  Gaynor, Mostashari, and Ginsburg \emph{Making Health Care Markets
  Work}.
\item
  Homework 3 due \textbf{Friday, 4/21}
\end{itemize}

\hypertarget{final-projects-due}{%
\subsubsection{4/24: Final projects due}\label{final-projects-due}}




\end{document}

\makeatletter
\def\@maketitle{%
  \newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
  \let \footnote \thanks
    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
}
%\fi
\makeatother