---
title: "Module 1: Health Insurance"
subtitle: "Part 1: Risk and Uncertainty"
author: Ian McCarthy | Emory University
date: Econ 372 #"`r format(Sys.time(), '%d %B %Y')`"
header-includes: 
  - \usepackage{tikz}
  - \usepackage{amsmath}
  - \usepackage{pgf}
  - \usepackage[xcolor=pst]{pstricks}
  - \usepackage{pgfshade}
  - \usepackage{pdftools}
output:
#  html_document: default  
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "macros.js"
knit: pagedown::chrome_print      
---

<!-- Adjust some CSS code for font size and maintain R code font size -->
<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 2em 1em 2em;    
}
.remark-code, .remark-inline-code { 
    font-size: 20px;
}
</style>


<!-- Set R options for how code chunks are displayed and load packages -->
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=3, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )

knitr::opts_hooks$set(fig.callout = function(options) {
  if(options$fig.callout) {
    options$echo = FALSE
  }
  options
})

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes,
               scales, gganimate, gapminder, gifski, png, tufte, plotly, OECD,
               ggrepel, survey, foreign, devtools, pdftools, xaringanExtra)
```


```{r xaringan-scribble, echo=FALSE}
xaringanExtra::use_scribble(rgb(0.9, 0.5, 0.5))
```


# Describing risk
We need three things to define risk in this class:
1. Probability
2. Expected value
3. Preferences (i.e., a utility function)

---
# 1. Probability
Definition: The likelihood that a given outcome will occur.<br>

--

Important to note the timing here...probability applies to an uncertain event that may have several possible outcomes. For example, I may have a heart attack or I may not. [Risk Calculator](http://www.cvriskcalculator.com/).

---
# 2. Expected value
Definition: The probability weighted average of the payoffs (or costs) associated with all possible outcomes.<br>

--

For two potential outcomes, $x_{1}$ and $x_{2}$, with probabilities $p_{1}$ and $p_{2}$:<br>

$E[x] = p_{1}x_{1} + p_{2}x_{2}$

---
# Example
What is my expected cost?
- Two possible outcomes: heart attack or no heart attack
- 10% chance of having a heart attack
- Cost of <span>&#36;</span>100,000 if I have a heart attack (but I will survive and recover)<br>

???
Give time for class to do this and answer. Type your answers in the chat once you have them.

---
# Answer

I will incur a cost of <span>&#36;</span>100,000 with 10% probability. So my expected cost is just $E[cost]=0.1*100,000 =$ `r format(0.1*100000,digits=5,big.mark=",")`.

---
# 3. Preferences
Definition: Preferences take the form of a utility function, $u(x)$, which tells us how much we benefit get from some consumption bundle, $x$.<br>

--

Expected utility combines expected value and utility...<br>
$E[u(x)] = p_{1}u(x_{1}) + p_{2}u(x_{2})$


???
Before moving on, this is a good time to pause and see how we're doing. These may seem like very simple concepts, but they are important. On our scale Zoom poll, "I understand the concepts of probability, expected value, and utility preferences."


---
# Risk preferences
With probabilities, expected values, and utilities/preferences, we can now measure preferences toward risk. <br>

--

- <b>Risk averse:</b> We prefer to avoid the risky situation. You would rather have the same (or slightly less) with certainty than a lottery over two risky outcomes.

- <b>Risk neutral:</b> Indifferent between the risky situation or that of certainty.

- <b>Risk loving:</b> Prefer the risky situation. 


???
Fun example...would you take the following bet: I'll give you 10 dollars with certainty, or we can flip a coin. Heads you get 15, tails you get 6.

Now let's try that again with a few more zeros. I'll give you 10 million, or we can flip a coin. Heads you get 15 million, tails you get 6 million.

The point: preferences toward risk are complicated! Luckily we don't really need to measure risk preferences to get the intuition we want...we just need to assume risk aversion.

---
# Risk aversion
Most common assumption is that individuals are risk averse. Mathematically, this follows from diminishing marginal utility.<br>

--

$u'(x_{1}) > u'(x_{2})$ for $x_{1} < x_{2}$


What does this mean in words?

---
# Risk aversion
```{tikz echo=FALSE, fig.ext='png', cache=TRUE, tikz-aversion}
\usetikzlibrary{calc}
\begin{tikzpicture}[my plot/.style={
                        thick,
                        smooth,
                        samples=100,
                        domain=0.1:5},
                    scale=0.75,
                    my grid/.style={dashed,opacity=0.5, every node/.style={black,opacity=1, transform shape}},
                    my axis/.style={latex-latex}]
\draw[my plot] (0,0) plot (\x,{ln(\x)});
\coordinate (start plot) at (0.1,{ln(0.1)}); % domain start
\coordinate (end plot) at (5,{ln(5)}); % domain end
%\draw[my axis] ([shift={(-0.5cm,0.5cm)}]start plot |- end plot) node[left] {$u(\cdot)$} |- node[coordinate](origin){} ([shift={(0.5cm,-0.5cm)}]start plot -| end plot) node[below] {$\cdot$};
\draw[my axis] ([shift={(-0.5cm,0.5cm)}]start plot |- end plot) |- node[coordinate](origin){} ([shift={(0.5cm,-0.5cm)}]start plot -| end plot);
\def\x{0.5}\def\y{4}\def\p{0.55} % define the x, y and p values
\coordinate (Ux) at (\x,{ln(\x)}); % set the u(x) coordinate
\coordinate (Uy) at (\y,{ln(\y)}); % set the u(y) coordinate
\coordinate (Up) at ({\p*\x+(1-\p)*\y},{ln(\p*\x+(1-\p)*\y)}); % set the u(p*x+(1-p)*y) coordinate
\draw (Ux) -- coordinate[pos=1-\p] (Up-mid) (Uy); % set the coordinate on the linear curve
\path let \p1=(Up-mid), \n1={pow(e,\y1*0.03514)} in (28.4576*\n1,\y1) coordinate (Up-mid2); 
\draw[my grid] (Ux) |- node[below]{$x_{1}$} (origin) |- node[left]{$u(x_{1})$} cycle;
\draw[my grid] (Uy) |- node[below]{$x_{2}$} (origin) |- node[left]{$u(x_{2})$} cycle;
\draw[my grid] (Up) |- node[below,font=\scriptsize]{$p_{1}x_{1}+p_{2}x_{2}$} (origin) |- node[left,font=\scriptsize]{$u(p_{1}x_{1}+p_{2}x_{2})$} cycle;
%\draw[my grid] (Up-mid2) |- node[below,font=\scriptsize]{$c(F,u)$} (origin) |- node[left]{$pu(x)+(1-p)u(y)$} cycle;
%\draw[my grid] (Up-mid) -- (Up-mid2);
\end{tikzpicture}
```


???
This graph has two different things on it. Let's focus on one at at time. Don't forget to use the pen!

---
# In-class Problem: Expected values
An individual starts with a wealth of <span>&#36;</span>100,000. With probability 0.3, they will get sick and incur a cost of <span>&#36;</span>40,000.

1. What is this person's expected cost of illness?
2. Assume this individual has a utility function of the form, $u(w) = w^{0.20}$. What is this person's expected utility?
3. Calculate this person's utility if they were to incur the expected cost of illness. Is this utility higher or lower than what you found in part (2)?



---
# Why purchase health insurance?
Say your utility function is $u(w)=\sqrt{w}$ and that you're starting with $w=$ <span>&#36;</span>100. I propose a lottery in which I flip a coin...heads you win <span>&#36;</span>20 and tails you lose <span>&#36;</span>20. <br>

--

1. What is the expected monetary value of this lottery? 

2. What is your utility at this expected value? 

3. What is the expected utility from this lottery?

---
# Answer
```{r include=FALSE}
ew <- .5*80+.5*120
eu <- .5*sqrt(80) + .5*sqrt(120)

```
Expected wealth is simply $\frac{1}{2} \times 80 + \frac{1}{2} \times 120 =$ `r format(ew,digits=3)`, which yields a utility of $u(w)=$ `r format(sqrt(ew),digits=3)`. But your expected utility is $\frac{1}{2} \times u(w_{heads}) + \frac{1}{2} \times u(w_{tails}) = \frac{1}{2} \times \sqrt{80} + \frac{1}{2} \times \sqrt{120} =$ `r format(eu, digits=3)`.


???
Put these values on our graph to see where we're headed.

Quick poll: Are we comfortable with the difference between expected utility and the utility at the expected value?

Takeaway so far - when you purchase health insurance, you do two things: 1) "pre-pay" for health care; and 2) pay to avoid financial risk.

