---
title: "Module 1: Health Insurance"
subtitle: "Demand for health insurance"
author: Ian McCarthy | Emory University
date: Econ 372 #"`r format(Sys.time(), '%d %B %Y')`"
header-includes: 
  - \usepackage{tikz}
  - \usepackage{amsmath}
  - \usepackage{pgf}
  - \usepackage[xcolor=pst]{pstricks}
  - \usepackage{pgfshade}
  - \usepackage{pdftools}
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "macros.js"
knit: pagedown::chrome_print
---

<!-- Adjust some CSS code for font size and maintain R code font size -->
<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 2em 1em 2em;    
}
.remark-code, .remark-inline-code { 
    font-size: 20px;
}
</style>


<!-- Set R options for how code chunks are displayed and load packages -->
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=3, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )

knitr::opts_hooks$set(fig.callout = function(options) {
  if(options$fig.callout) {
    options$echo = FALSE
  }
  options
})

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes,
               scales, gganimate, gapminder, gifski, png, tufte, plotly, OECD,
               ggrepel, survey, foreign, devtools, pdftools)
```

# Quick summary

Let's quickly revisit the main things from last class:

1. Motivating story: Humana leaving the exchanges due to poor risk pool.
2. Preliminary things: Calculate expected values and expected utilities

???
Revisit the expected utility calculation from the end of last class and do the problem together.

Note the weekly check-ins.

---
# Why purchase health insurance?
Say your utility function is $u(w)=\sqrt{w}$ and that you're starting with $w=$ <span>&#36;</span>100. I propose a lottery in which I flip a coin...heads you win <span>&#36;</span>20 and tails you lose <span>&#36;</span>20. <br>

--

1. What is the expected monetary value of this lottery? 

2. What is your utility at this expected value? 

3. What is the expected utility from this lottery?

---
# Answer
```{r include=FALSE}
ew <- .5*80+.5*120
eu <- .5*sqrt(80) + .5*sqrt(120)

```
Expected wealth is simply $\frac{1}{2} \times 80 + \frac{1}{2} \times 120 =$ `r format(ew,digits=3)`, which yields a utility of $u(w)=$ `r format(sqrt(ew),digits=3)`. But your expected utility is $\frac{1}{2} \times u(w_{heads}) + \frac{1}{2} \times u(w_{tails}) = \frac{1}{2} \times \sqrt{80} + \frac{1}{2} \times \sqrt{120} =$ `r format(eu, digits=3)`.


???
Put these values on our graph to see where we're headed.

Quick poll: Are we comfortable with the difference between expected utility and the utility at the expected value?

---
# Risk premium

The maximum amount of money that a risk-averse person is willing to pay to avoid the risky scenario. In other words, the amount of money that makes a person indifferent between the certain and uncertain situations.<br>

--

This is part of how we measure willingness-to-pay for health insurance.

---
# Risk premium
```{tikz echo=FALSE, fig.ext='png', cache=TRUE, tikz-premium}
\usetikzlibrary{calc}
\begin{tikzpicture}[my plot/.style={
                        thick,
                        smooth,
                        samples=100,
                        domain=0.1:5},
                    scale=0.75,
                    my grid/.style={dashed,opacity=0.5, every node/.style={black,opacity=1}, transform shape},
                    my axis/.style={latex-latex}]
\draw[my plot] (0,0) plot (\x,{ln(\x)});
\coordinate (start plot) at (0.1,{ln(0.1)}); % domain start
\coordinate (end plot) at (5,{ln(5)}); % domain end
\draw[my axis] ([shift={(-0.5cm,0.5cm)}]start plot |- end plot) |- node[coordinate](origin){} ([shift={(0.5cm,-0.5cm)}]start plot -| end plot);
\def\x{0.5}\def\y{4}\def\p{0.55} % define the x, y and p values
\coordinate (Ux) at (\x,{ln(\x)}); % set the u(x) coordinate
\coordinate (Uy) at (\y,{ln(\y)}); % set the u(y) coordinate
\coordinate (Up) at ({\p*\x+(1-\p)*\y},{ln(\p*\x+(1-\p)*\y)}); % set the u(p*x+(1-p)*y) coordinate
\draw (Ux) -- coordinate[pos=1-\p] (Up-mid) (Uy); % set the coordinate on the linear curve
\path let \p1=(Up-mid), \n1={pow(e,\y1*0.03514)} in (28.4576*\n1,\y1) coordinate (Up-mid2); 
\draw[my grid] (Ux) |- node[below]{$x_{1}$} (origin) |- node[left]{$u(x_{1})$} cycle;
\draw[my grid] (Uy) |- node[below]{$x_{2}$} (origin) |- node[left]{$u(x_{2})$} cycle;
%\draw[my grid] (Up) |- (origin) |- cycle;
\draw[my grid] (Up) |- (origin);
%\draw[my grid] (Up-mid2) |- node[below,font=\scriptsize]{$c(F,u)$} (origin) |- node[left]{$pu(x)+(1-p)u(y)$} cycle;
\draw[my grid] (Up-mid2) |- node[below right]{\hspace{.05cm} $\pi$} (origin) |- node[left,font=\scriptsize]{$E[u(x)]$} cycle;
\draw[my grid] (Up-mid) -- (Up-mid2);
\end{tikzpicture}
```

???
This is where we'll end for the first class, 2/4. Now let's go through a quick in-class quiz.

End class with discussion/preview of homework 1 and opportunity to ask questions regarding the final project or other assignments.

---
# Example
Consider the utility function, $u(w)=\ln(w)$. An individual starts with a wealth of <span>&#36;</span>100,000. With probability 0.25, this person will get sick and incur a cost of <span>&#36;</span>20,000. Their wealth in the sick state is therefore <span>&#36;</span>80,000. What is the maximum amount this person is willing to pay for health insurance?<br>

--

1. Calculate expected wealth, $E[w]$.
2. Calculate expected utility, $E[U(w)]$.
3. Calculate value of wealth that gives you $u=E[U(w)]$ (based on the utility function).
4. Calculate the risk premium as the difference between (1) and (3).
5. Calculate maximum willingness to pay by adding the risk premium and the expected cost.

---
# Answer
```{r, include=FALSE}
ecost <- .25*20000
eu <- .75*log(100000) + .25*log(80000)
inv.u <- exp(eu)
wtp <- 100000-inv.u
```
We're asked to find some wealth level, $y$, such that the person is indifferent between $y$ with certainty versus the risky wealth levels, $w_{h}=$ <span>&#36;</span>100,000 with probability 0.75 or $w_{s}=$ <span>&#36;</span>80,000 with probability 0.25.<br>

The person's expected utility with uncertainty is: $E[u]=0.75\times \ln (100000) + 0.25 \times \ln (80000) =$ `r eu`. We therefore need to find $y$ such that $u(y)=$ `r eu`. Given our utility function, this is satisfied for $y=$ <span>&#36;</span>`r format(inv.u,digits=5,big.mark=",")`. Since the person starts with <span>&#36;</span>100,000, they are willing to pay a maximum of <span>&#36;</span>100,000 $-$ <span>&#36;</span>`r format(inv.u,digits=5,big.mark=",")` $=$ <span>&#36;</span>`r format(wtp,digits=5,big.mark=",")` for health insurance.

Finally, since the expected cost of care is <span>&#36;</span>`r format(ecost,digits=5,big.mark=",")`, we can break this <span>&#36;</span>`r format(wtp,digits=5,big.mark=",")` into its actuarily fair premium of <span>&#36;</span>`r format(ecost,digits=5,big.mark=",")` plus the loading factor or risk premium of <span>&#36;</span>`r format(wtp-ecost,digits=5,big.mark=",")`.

---
# In-class Problem: Demand for insurance
Assume that utility takes the log form, $u(x)=ln(x)$. If someone is healthy, they maintain their current wealth of <span>&#36;</span>100, and if they become ill, they must incur a cost of <span>&#36;</span>50. Answer the following questions based on this setup.

1. Calculate the risk premium and willingness to pay based on a probability of illness of 0.1.
2. Repeat part (1) using a probability of illness of 0.2.
3. Repeat part (1) using a probability of illness of 0.5.
4. Explain how these values differ and why. What might this say about the profitability of insurance in a market with many sick people?


---
# What affects the risk premium?

Based on the graph, what do you think are some things that might affect the risk premium?<br>

--

1. Curvature of the utility function

2. Probability of illness

3. Cost of illness


---
# High risk pools

- A "high-risk pool" is a way to put people that are more likely to incur high medical costs all in one plan. 
- Recalling the curvature of demand function, probability of illness, and cost of illness...do you think a high-risk pool is sustainable (think about the profit to the insurer)?<br>

--

Let's look at this in practice, [KFF High-risk Pools](https://www.kff.org/health-reform/issue-brief/high-risk-pools-for-uninsurable-individuals/)


---
# Other reasons to buy health insurance

1. Increase bargaining power with providers

2. Manage where care is delivered (due to information problems in health care decisions)


---
# Main takeaways

1. Explain and show graphically how changes in the utility function, probability of illness, and the cost of illness affect the risk premium

2. Calculate amount of insurance individuals would purchase under given assumptions

