---
title: "Day 1: Morning Supplement"
subtitle: "a quick survey to start us off"
author: Ian McCarthy | Emory University
date: Econ 372 #"`r format(Sys.time(), '%d %B %Y')`"
header-includes: 
  - \usepackage{graphicx}
  - \usepackage{amsmath}
output:
#  html_document: default (toggle on for "simplified" view)
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "macros.js"
---
count: false
class: clear

<!-- Adjust some CSS code for font size, maintain R code font size -->
<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 2em 1em 2em;    
}
.remark-code, .remark-inline-code { 
    font-size: 20px;
}
</style>


<!-- Set R options for how code chunks are displayed and load packages -->
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(dplyr.summarise.inform = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=3, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T#, echo=F, warning=F, message=F
  )

knitr::opts_hooks$set(fig.callout = function(options) {
  if(options$fig.callout) {
    options$echo = FALSE
  }
  options
})

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes,
               scales, gganimate, gapminder, gifski, png, tufte, plotly, OECD,
               ggrepel, xaringanExtra)
```


```{r xaringan-scribble, echo=FALSE}
xaringanExtra::use_scribble(rgb(0.9, 0.5, 0.5))
```


**Among americans with employer-provided health insurance, what proportion accounts for about a fifth of the total medical spending?**

.pull-left[
  A.  12%
  
  B.  10%
  
  C.  6%
  
  D.  1%
]


---
count: false
class: clear

**Among americans with employer-provided health insurance, what proportion accounts for about a fifth of the total medical spending?**

.pull-left[
  A.  12%
  
  B.  10%
  
  C.  6%
  
  **D.  1%**
]

.pull-right[
A Kaiser Family Foundation survey found that 1.3% of people with health insurance through their employers accounted for 19.5% of spending (about $88,000 per person on average)
]

---
class: clear

**In 2020, what was the average cost of employer-provided health insurance for a family?**

.pull-left[
  A.  $16,450
  
  B.  $35,200
  
  C.  $21,350

  D.  $12,500
]


---
count: false
class: clear

**In 2020, what was the average cost of employer-provided health insurance for a family?**

.pull-left[
  A.  $16,450
  
  B.  $35,200
  
  **C.  $21,350**

  D.  $12,500
]

.pull-right[
KFF employer survey - average worker contribution was about $5,600
]


---
class: clear

**Not counting blood tests, drugs, or other treatments, the average cost of an ER visit in 2017 was $1,390. How much did this increase over the prior decade?**

.pull-left[
  A.  0.76%
  
  B.  7.6%
  
  C.  76%
  
  D.  176%
]


---
count: false
class: clear

**Not counting blood tests, drugs, or other treatments, the average cost of an ER visit in 2017 was $1,390. How much did this increase over the prior decade?**

.pull-left[
  A.  0.76%
  
  B.  7.6%
  
  C.  76%
  
  **D.  176%**
]

.pull-right[
Data from the Health Care Cost Institute show a cost of $503 in 2008.
]



---
class: clear

**What is the largest category of spending on health care in the U.S.?**

.pull-left[
  A.  Hospital services
  
  B.  Physician services
  
  C.  Prescription drugs
  
  D.  Nursing homes
  
  E.  Waste
]


---
count: false
class: clear

**What is the largest category of spending on health care in the U.S.?**

.pull-left[
  **A.  Hospital services**
  
  B.  Physician services
  
  C.  Prescription drugs
  
  D.  Nursing homes
  
  E.  Waste
]

.pull-right[
From CMS National Health Expenditure reports, we spent $1.2 trillion on hospital care in 2019 (about 31% of total spending). Prescription drugs, while growing at a faster rate, constitute less than 10% of total health expenditures.
]


<!-- New Section -->
---
class: inverse, center, middle

# What are we doing today?

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1055px></html>

---
# Goals

- Motivate the study of health economics
- Review the class, assignments, etc.
- Become familiar with class website and Canvas page



<!-- New Section -->
---
class: inverse, center, middle

# Looking ahead

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1055px></html>


---
# Outside of class

- Review the math practice problems on the class website
- Nothing to turn in